 

### 2021-06-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Wind speeds are also higher at the top of the canopy, owing to greater exposure to macroclimatic conditions with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
Leaf trait comparison within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation.

Therefore, across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).

@@

Within-canopy leaf traits vary along the canopy light gradient, ie. in sun and shade leaves [@casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. This variation is crucial to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity. Further, research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

@@

Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater , hydraulically efficiency than shade leaves [Brodribb et al. 2010]. All these traits, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous canopy light conditions, sun-leaves in different canopy orientations are primarily shaped by differences in $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a]. Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [PW West, 2019]. This difference can alter tree growth patterns in response to biophysical conditions. Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

@@

###warming
Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [Keenan et al. 2019]. However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pamela templer et al.].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].

>>>>

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
Leaf trait variation within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation. Therefore, across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).

@@

Within-canopy leaf traits vary along the canopy light gradient, ie. in sun and shade leaves [@casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. This variation is crucial, to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity. Further, research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

@@

Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater hydraulically efficiency than shade leaves [Brodribb et al. 2010]. All these traits, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous canopy light conditions, sun-leaves in different canopy orientations are primarily shaped by differences in $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a]. Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [PW West, 2019]. This difference can alter tree growth patterns in response to biophysical conditions. Particularly, higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

@@

### warming

Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [Keenan et al. 2019]. However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pamela templer et al.].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
